Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.tex
Section 3 highlighting

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we
show\footnote{\textcolor{violet}{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme). I'm happy with other synonyms as well.} \textcolor{orange}{accept}}
how our package's functions fit into the data analysis pipeline
introduced in Fig. \ref{fig:pipeline}. All front-facing functions in
\pkg{EpiCompare} are aimed to be as user-friendly as possible. We also
focus on providing the user ``tidyverse'' style functions, that
encourage piping objects from one function to the next and follow clear
``verb'' naming schemes \citep{Wickham2019}. Although users can
incorporate \pkg{EpiCompare} into any step in the data analysis
pipeline, there are two primary points of entry. The first point of
entry is the very beginning with pre-processing and visualizing raw
data, and the second point of entry is after modeling and simulation.
Figure \ref{fig:pipeline2} captures these different paths, and we
highlight how to leverage \pkg{EpiCompare} functionalities in the
subsections below.

@@

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
\sout{Raw data can appear in many different formats.}\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\textcolor{orange}{I think we can get rid of the 'raw data can appear...' sentence.} RESOLVED 5/17}
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. Compartment totals (amounts of individuals in each state) are
then imputed from those case counts other information about the disease
and the population of interest. In \pkg{EpiCompare}, we focus on
understanding the overall impact of an outbreak at the
aggregate/population level, which allows for streamlined examination of
overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
There are often situations where
\textcolor{violet}{\sout{aggregating} grouping}\footnote{RESOLVED 5/17 (aggregating)}
agents into subpopulations (e.g.~subpopulations defined by age or sex)
can highlight different aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for differing disease dynamics in different subpopulations.
In \pkg{EpiCompare}, we facilitate subpopulation analysis by extending
\texttt{agents\_to\_aggregate()} to work with
\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
split-apply-combine paradigm for data analysis
\citep{wickham2011}.\textbackslash footnote\{Changed to try to clarify
how how agents\_to\_aggregate works with group\_by and why
(split-apply-combine)\}

\textcolor{orange}{old text below}

{[}by combining the function \pkg{dplyr}::\texttt{group\_by()} and
\texttt{agents\_to\_aggregate()} to provide aggregation at a
subpopulation
level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}

Besides aiding subpopulation analysis, \texttt{agents\_to\_aggregate()}
can
accommodate\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc. Again, I'm open to other synonyms.]}\textcolor{orange}{I like accomodate}  RESOLVED 5/17}
a wide range of information about each individual. In fact, this
function can account for infinitely many states. This functionality
allows the practitioner to aggregate information relative to common
states (e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well
as more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
death/exit and birth/entry dates. Overall, this function is a powerful
tool for pre-processing data.

**Exploratory Data Analysis (EDA)*

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there \textcolor{violet}{are} three
unique epidemiological states (like in the SIR model). Building on
\pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s
\texttt{geom\_aggregate()} provides a
\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \textcolor{orange}{accepted}. RESOLVED 5/17}
way to explore how different subpopulations experience of an epidemic
\citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because (1) SIR models are some of the most common
and basic epidemic state-based models and (2) our three-dimensional
simplex representation of these epidemics emphasizes a
time-invariant\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \textcolor{orange}{agreed} RESOLVED 5/17}
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is
often\footnote{\textcolor{violet}{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence. Maybe it's just me.]}RESOLVED 5/17}
model fitting and/or simulation. In this step and the next step
(post-processing), we highlight how practitioners can pair models and
simulations of epidemics from outside of \pkg{EpiCompare} with analysis
and simulation tools in \pkg{EpiCompare}. While \pkg{EpiCompare} does
not focus on fitting models to data, we do provide some flexible
functions for simulation of basic discrete-time epidemic-state models.
These functions simulate individual-level information based on
practitioner estimated transition rates between states and can be
combined with \texttt{agents\_to\_aggregate()} to view these simulations
through an aggregate lens. The function \texttt{simulate\_SIR\_agents()}
simulates a basic SIR epidemic with user inputs for the number of
simulations, the initial number in each state, the infection and
recovery parameters \((\beta, \gamma\)), and the total number of
discrete time steps. This function allows for easy access to SIR model
analysis and comparison. Beyond SIR models, the function
\texttt{simulate\_agents()} takes as input a user-specified
state-transition matrix and other epidemic parameters to allow the user
to create simulations for an outbreak with \textit{any} number of states
and any number of transitions among them. This flexibility in states can
be used to also reflect group-based dynamics. In turn, this allows for
users to explore the space of models in an intuitive way without getting
bogged down by too much mathematical detail. For consistency, we have
made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.

@@

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format. In general,
post-processing of modeling and simulation consists of making summary
statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
functions to standardize simulation and model output from external
packages and (2) a function to transform standardized simulation and
model output into a format amenable to time-invariant
analysis.\footnote{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}

Modeling and simulation output can be very complex objects, and as a
result, a number of epidemic modeling \proglang{R} packages return a
special class. The special classes often contain a plethora of
information about residuals, model diagnostics, input parameters, and
more. While incredibly useful, these special classes can be difficult
for novice coders to handle.
\textcolor{orange}{motivates for fortify but not filamental compression}.
To this end, \pkg{EpiCompare} provides a series of
\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
which transform output from infectious disease modeling and simulation
packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
which contain information about the total number of individuals in each
state at a given time, for a given simulation. These fortify functions
have output that is consistent with that of
\texttt{agents\_to\_aggregate()}. These standardized outputs can then be
piped to summaries, tables, and plots.

Because the fortified output contains temporal information and we
provide the function, \texttt{filament\_compression()}, to transform
temporally defined epidemics to their filamental representations. These
filaments can then be fairly compared to one another or passed to
further time-invariant analysis tools described below.

{[}{[}\textcolor{orange}{OLD TEXT BELOW}

@@

\textbf{Comparisons and Assessment}

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare different models, and
simulations to one another. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for model and simulation results that
extend beyond the standard performance metrics (e.g.~mean squared error
or AIC) and into the lens of time-invariant
analysis.\footnote{a bit of an awk transition but I wanted to use past wording about TIA.}
We have found these tools to be specifically applicable for situations
where only one ``season'' or ``cycle'' of an epidemic has occurred.

@@

{[}{[}\textcolor{orange}{OLD TEXT}

@@

Finally, in this section we show how tools from \pkg{EpiCompare} can be
used in each step of the data analysis pipeline shown in Fig.
\ref{fig:pipeline}. We analyze an outbreak of measles in the town of
Hagelloch, Germany from 1861-1862, a data set organized by
\cite{pfeilsticker1863}. The data was later made visible by
\cite{oesterle1992} and made available in an \proglang{R} by
\cite{surveillance2017}. This data set includes a rich collection of
features and is an ideal testing ground for methodology in infectious
disease epidemiology
\cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\textcolor{orange}{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's goingon.}}

{[}{[}\textcolor{orange}{OLD TEXT}

>>>>

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we show how our package's functions fit into the
data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions in \pkg{EpiCompare} are aimed to be as
user-friendly as possible. We also focus on providing the user
``tidyverse'' style functions, that encourage piping objects from one
function to the next and follow clear ``verb'' naming schemes
\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
into any step in the data analysis pipeline, there are two primary
points of entry. The first point of entry is the very beginning with
pre-processing and visualizing raw data, and the second point of entry
is after modeling and simulation. Figure \ref{fig:pipeline2} captures
these different paths, and we highlight how to leverage \pkg{EpiCompare}
functionalities in the subsections below.

@@

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. Compartment totals (amounts of individuals in each state) are
then imputed from those case counts other information about the disease
and the population of interest. In \pkg{EpiCompare}, we focus on
understanding the overall impact of an outbreak at the
aggregate/population level, which allows for streamlined examination of
overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for differing disease dynamics in different subpopulations.
In \pkg{EpiCompare}, we facilitate subpopulation
analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
by extending \texttt{agents\_to\_aggregate()} to work with
\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
split-apply-combine paradigm for data analysis \citep{wickham2011}.

{[}{[}\textcolor{red}{OLD SENTENCE} by combining the function
\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()}
to provide aggregation at a subpopulation
level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}{]}

Besides aiding subpopulation analysis, \texttt{agents\_to\_aggregate()}
can accommodate a wide range of information about each individual. In
fact, this function can account for infinitely many states. This
functionality allows the practitioner to aggregate information relative
to common states (e.g.~``Susceptible'', ``Infectious'', and
``Recovered'') as well as more complex states (e.g.~``Exposed'',
``iMmune'', ``Hospitalized''). Additionally,
\texttt{agents\_to\_aggregate()} permits indicators for death/exit and
birth/entry dates. Overall, this function is a powerful tool for
pre-processing data.

\textbf{Exploratory Data Analysis (EDA)}

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there are three unique epidemiological
states (like in the SIR model). Building on \pkg{ggplot2} and
\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
provides a way to explore how different subpopulations experience of an
epidemic \citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because (1) SIR models are some of the most common
and basic epidemic state-based models and (2) our three-dimensional
simplex representation of these epidemics emphasizes a time-invariant
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

@@

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is often model fitting
and/or simulation. In this step and the next step (post-processing), we
highlight how practitioners can pair models and simulations of epidemics
from outside of \pkg{EpiCompare} with analysis and simulation tools in
\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
models to data, we do provide some flexible functions for simulation of
basic discrete-time epidemic-state models. These functions simulate
individual-level information based on practitioner estimated transition
rates between states and can be combined with
\texttt{agents\_to\_aggregate()} to view these simulations through an
aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
a basic SIR epidemic with user inputs for the number of simulations, the
initial number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified state-transition matrix and other epidemic
parameters to allow the user to create simulations for an outbreak with
\textit{any} number of states and any number of transitions among them.
This flexibility in states can be used to also reflect group-based
dynamics. In turn, this allows for users to explore the space of models
in an intuitive way without getting bogged down by too much mathematical
detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.

@@

\textcolor{orange}{[[NEW TEXT}

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format. In general,
post-processing of modeling and simulation consists of making summary
statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
functions to standardize simulation and model output from external
packages and (2) a function to transform standardized simulation and
model output into a format amenable to time-invariant
analysis.\footnote{\textcolor{orange}{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

Modeling and simulation output can be very complex objects, and as a
result, a number of epidemic modeling \proglang{R} packages return a
special class. The special classes often contain a plethora of
information about residuals, model diagnostics, input parameters, and
more. While incredibly useful, these special classes can be difficult
for novice coders to handle. To this end, \pkg{EpiCompare} provides a
series of fortify-style methods, called \code{fortify_aggregate()} which
transform output from infectious disease modeling and simulation
packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
which contain information about the total number of individuals in each
state at a given time, for a given simulation. These fortify functions
have output that is consistent with that of
\code{agents_to_aggregate()}. These standardized outputs can then be
piped to summaries, tables, and plots.

Because the fortified output contains temporal information and we
provide the function, \code{filament_compression()}, to transform
temporally defined epidemics to their filamental representations. These
filaments can then be fairly compared to one another or passed to
further time-invariant analysis tools described below.
\textcolor{orange}{]]}

{[}{[}\textcolor{red}{OLD TEXT BELOW}

@@

\textbf{Comparisons and Assessment} \textcolor{orange}{[[ NEW TEXT}

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare different models, and
simulations to one another. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for model and simulation results that
extend beyond the standard performance metrics (e.g.~mean squared error
or AIC) and into the lens of time-invariant
analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
We have found these tools to be specifically applicable for situations
where only one ``season'' or ``cycle'' of an epidemic has occurred.

@@

\textcolor{orange}{]]}

{[}{[}\textcolor{red}{OLD TEXT}

@@

\textcolor{orange}{[[NEW TEXT}

Finally, in this section we show how tools from \pkg{EpiCompare} can be
used in each step of the data analysis pipeline shown in Fig.
\ref{fig:pipeline}. We analyze an outbreak of measles in the town of
Hagelloch, Germany from 1861-1862, a data set organized by
\cite{pfeilsticker1863}. The data was later made visible by
\cite{oesterle1992} and made available in an \proglang{R} by
\cite{surveillance2017}. This data set includes a rich collection of
features and is an ideal testing ground for methodology in infectious
disease epidemiology
\cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\textcolor{orange}{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's going on.}}

\textcolor{orange}{]]}

{[}{[}\textcolor{red}{OLD TEXT}

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.tex
Section 3 highlighting

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (where
\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.

>>>>

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (where
\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'\,' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.
